---
 title: "Mycobiome statistical analysis"
 author: 
    - name: "Rachelle Fernández-Vargas"
    affiliation: "University of Costa Rica"
    - name: "Keilor Rojas-Jiménez"
    affiliation: "University of Costa Rica"
 description: |
   Statistical analysis and visualization of ITS metabarcoding data 
   including alpha diversity, NMDS, PERMANOVA and Venn diagram
inputs:
   - 3biom-Mycobiome.xlsx (raw data)
   - Venn.xlsx
 outputs:
   - Figure1-RelAbundOrder-nod.png
   - Figure2a-Diversity.pdf
   - Figure2b-Diversity.png
   - Figure3-NMDS.png
   - venn_diagram.png

 repository: "https://github.com/vrachfer00/MycobiomeAnalysis.git"
---

Set working directory to the location of the result files (update to match your data location)
```{r}
setwd("D:/Sequences/...")

# Create a new directory for storing extracted ITS data (if not already created)
dir.create("1Extraction-ITS")

```

Load required libraries for data manipulation, visualization, and analysis
```{r}
library(tidyr)
library(readxl)
library(ggplot2)
library(dplyr)
library(vegan)
library(RColorBrewer)
```

Load the dataset from the Excel file
```{r}
#The "RelAbundOrder" sheet contains relative abundance data for fungal orders
data <- read_excel("3biom-Mycobiome.xlsx",sheet= "RelAbundOrder")%>% 
  pivot_longer(cols =   Pleosporales:Other, 
               names_to = "Class", 
               values_to = "valor")
data

# Generate a color palette based on the number of unique fungal classes in the data
cols <- colorRampPalette(brewer.pal(12, "Paired"))
myPal <- cols(length(unique(data$Class)))
```

Create a stacked bar plot showing relative abundance of fungal classes per sample
```{r}
# Data is grouped by nodulation status
ggplot(data=data, aes(x=Sample, y=valor,fill=Class))+
  geom_bar(stat = "identity")+ 
  facet_wrap(~ Nodulation,scales="free_x")+
  scale_fill_manual(values = myPal)+
  theme_bw()+ylab("Relative Abundance") + xlab("")+  
  theme(plot.title = element_text(hjust = 0,size = 8, face="bold"),
        axis.text.x = element_text(size=8,angle = 90,vjust = 0.5, hjust=1),
        axis.text.y = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 7),
        legend.position = "right",
        legend.key.size = unit(0.3, "cm"),
        legend.title = element_blank())
ggsave("Figure1-RelAbundOrder-nod.png",width = 17.5, height = 8.5, units = "cm",dpi = 300)
```


########## Diversity  ############
```{r}
#Load necessary libraries for data manipulation, visualization, and diversity analysis
library(ggplot2)
library(dplyr)
library(Hmisc)
library(tidyr)
library(vegan)
library(dplyr)
library(grid)
library(plotrix)
library(fossil)
```

Read environmental data from an Excel file (contains metadata for samples)
```{r}
env <- read_excel("3biom-Mycobiome.xlsx", sheet= "nmds")
env  # Display the environmental data
```

Read microbial diversity data from a CSV file
```{r}
div <- read.csv2("nmds.csv", header=T, sep=";", row.names=1)
```

Normalize abundance data using total sum scaling (each row sums to 1)
```{r}
div <- decostand(div, "total")
str(div)  # Check the structure of the transformed data
```

Calculate alpha diversity indices
```{r}
Richness <- rowSums(div > 0)       # Count of unique taxa (ASV richness)
Shannon <- diversity(div)          # Shannon entropy index (measures diversity)
Pielou <- Shannon/log(Richness)    # Pielou's evenness index (measures species evenness)
```

Create a dataframe to store calculated diversity indices
```{r}
diversity <- data.frame(Richness, Shannon, Pielou)
write.csv2(diversity, "div.csv")  # Save diversity data to a CSV file
```

Read processed diversity data from a CSV file
```{r}
diver <- read.csv2("div1.csv", header=T, sep=";")
str(diver)  # Check the structure of the dataframe
```


Create boxplots to visualize diversity metrics across species and nodulation status
```{r}
a1 = ggplot(diver, aes(factor(Species), Richness)) + 
  geom_boxplot() + facet_wrap(~ Nodulation, scales="free_x") +
  theme_bw() + ylab("ASV Richness") + xlab("") + 
  theme(axis.text.x = element_blank(), 
        axis.title.y = element_text(size=9, face="bold"),
        axis.ticks.x = element_blank())

a2 = ggplot(diver, aes(factor(Species), Shannon)) + 
  geom_boxplot() + facet_wrap(~ Nodulation, scales="free_x") +
  theme_bw() + ylab("Shannon Index") + xlab("") + 
  theme(axis.text.x = element_blank(), 
        axis.title.y = element_text(size=9, face="bold"),
        axis.ticks.x = element_blank())

a3 = ggplot(diver, aes(factor(Species), Pielou)) + 
  geom_boxplot() + facet_wrap(~ Nodulation, scales="free_x") +
  theme_bw() + ylab("Pielou") + xlab("") + 
  theme(axis.text.x = element_text(size=9, face="bold.italic", angle=90, vjust=0.5, hjust=1),
        axis.title.y = element_text(size=9, face="bold"),
        axis.ticks.x = element_blank())
```

Save the plots as a PDF file
```{r}
library(patchwork)  # For arranging multiple plots
pdf("Figure2a-Diversity.pdf", width=3.4, height=7)
a1 / a2 / a3  # Arrange plots vertically
dev.off()
```

Compare diversity metrics between nodulation categories
```{r}
b1 = ggplot(diver, aes(factor(Nodulation), Richness)) + 
  geom_boxplot() + geom_jitter(width = 0.1) + 
  geom_text(data = NULL, x = 1.5, y = 185, label = "P=0.0002*", size=3) +
  theme_bw() + ggtitle("A") + ylab("ASV Richness") + xlab("") + 
  theme(plot.title = element_text(size=9, face="bold"), 
        axis.text.x = element_blank(),
        axis.title.y = element_text(size=9, face="bold"),
        axis.ticks.x = element_blank())

b2 = ggplot(diver, aes(factor(Nodulation), Shannon)) + 
  geom_boxplot() + geom_jitter(width = 0.1) + 
  geom_text(data = NULL, x = 1.5, y = 3.5, label = "P=0.003*", size=3) +
  theme_bw() + ylab("Shannon Index") + xlab("") + 
  theme(axis.text.x = element_text(size=8, face="bold"),
        axis.title.y = element_text(size=9, face="bold"),
        axis.ticks.x = element_blank())
```

Perform Wilcoxon tests to compare diversity metrics between nodulation groups
```{r}
wilcox.test(Richness ~ Nodulation, data = diver)  # P=0.0002
wilcox.test(Shannon ~ Nodulation, data = diver)   # P=0.003
wilcox.test(Pielou ~ Nodulation, data = diver)    # P>0.05 (not significant)
```

Compare diversity metrics between plant subfamilies
```{r}
b4 = ggplot(diver, aes(factor(Subfamily), Richness)) + 
  geom_boxplot() + geom_jitter(width = 0.1) + 
  geom_text(data = NULL, x = 1.5, y = 185, label = "P>0.05", size=3) +
  theme_bw() + ggtitle("B") + ylab("") + xlab("") + 
  theme(plot.title = element_text(size=9, face="bold"),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size=9, face="bold"),
        axis.ticks.x = element_blank())

b5 = ggplot(diver, aes(factor(Subfamily), Shannon)) + 
  geom_boxplot() + geom_jitter(width = 0.1) + 
  geom_text(data = NULL, x = 1.5, y = 3.5, label = "P>0.05", size=3) +
  theme_bw() + ylab("") + xlab("") + 
  theme(axis.text.x = element_text(size=8, face="bold"),
        axis.title.y = element_text(size=9, face="bold"),
        axis.ticks.x = element_blank())
```

Perform Wilcoxon tests for subfamily comparisons
```{r}
wilcox.test(Richness ~ Subfamily, data = diver)  # P>0.05
wilcox.test(Shannon ~ Subfamily, data = diver)   # P>0.05
wilcox.test(Pielou ~ Subfamily, data = diver)    # P>0.05
```

Combine plots for final visualization
```{r}
library(patchwork)
(b1 + b4) / (b2 + b5)  # Arrange plots in a grid
```

Save the final figure as a PNG file
```{r}
ggsave("Figure2b-Diversity.png", width = 13, height = 15, units = "cm", dpi = 300)
```

############ NMDS ###############
```{r}
#Load necessary libraries 
library(ggplot2)
library(dplyr)
library(vegan)
library(Hmisc)
library(tidyr)
```

Read the species abundance data (community matrix)
```{r}
nmds <- read.csv2("nmdsorder.csv", header=TRUE, row.names=1)  # Load species data
str(nmds) 
```

Read the environmental data (metadata)
```{r}
env <- read.csv2("env.csv", header=TRUE, row.names=1)  # Load environmental variables
str(env)
```

Standardize the community matrix
```{r}
nmds <- decostand(nmds, "total")  # Standardization using total sum scaling
```

Perform Non-Metric Multidimensional Scaling (NMDS)
```{r}
nmds <- metaMDS(nmds, trace = FALSE)  # Run NMDS without verbose output
nmds  # Display the NMDS results
```

Extract NMDS coordinates and create a dataframe for plotting
```{r}
data.scores = as.data.frame(nmds$points)  # Convert NMDS scores to a dataframe
data.scores$Dominance <- rownames(data.scores)  # Add row names as a new column

grp <- env$Nodulation  # Extract the nodulation status from environmental data
data.scores$Nodulation <- grp  # Add nodulation status to the NMDS scores dataframe

data.scores  # View the processed NMDS scores
```

Generate NMDS plot
```{r}
#pdf("Figure3b-NMDS.pdf",width=5,height= 4)  # Optionally save as a PDF
ggplot(data = data.scores, aes(x = MDS1, y = MDS2)) + 
  geom_point(data = data.scores, aes(colour = Nodulation), size = 3, alpha = 0.5) + 
  stat_ellipse(aes(color = Nodulation)) +
  geom_text(data=data.scores,aes(x=MDS1,y=MDS2,label=Dominance),size=1.8,vjust=2) +
  geom_text(aes(0.25,1.5, label = "Permanova, P=0.004*"),size=2.5)+
  ggtitle("")+ ylab("NMDS2") + xlab("NMDS1")+
  theme(panel.background = element_blank(),
        panel.border = element_rect(fill = NA, colour = "grey30"), 
        plot.title = element_text(size = 9, face="bold"),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 8),
        axis.title.y = element_text(size = 8),
        legend.direction = "vertical",
        legend.text = element_text(size = 6), 
        legend.title = element_blank(),
        legend.key.size = unit(0.4, "cm"))

ggsave("Figure3-NMDS.png",width = 10, height = 10, units = "cm",dpi = 300)
```

###### PERMANOVA Analysis ######

```{r}
per <- read.csv2("nmdsorder.csv", header=TRUE, row.names=1)
per <- decostand(per, "total")  # Standardize data again

env <- read.csv2("env.csv", header=T)  # Reload environmental data
```


Perform Constrained Correspondence Analysis (CCA) and PERMANOVA
```{r}
# Test the effect of Nodulation on community composition
Nod <- cca(per ~ Nodulation, env)  # Perform CCA
anova(Nod, step=1000)  # Test significance of Nodulation effect
adonis2(per ~ Nodulation, env, perm=1000)  # Run PERMANOVA (P=0.004, significant effect)
```


Test the effect of Subfamily on community composition
```{r}
Sub <- cca(per ~ Subfamily, env)
anova(Sub, step=1000)
adonis2(per ~ Subfamily, env, perm=1000)  # P > 0.05, no significant effect
```


#############VENN######################

```{r}
# Install necessary packages 
install.packages("ggvenn")    
install.packages("RColorBrewer") 
# Load required libraries
library(ggvenn)        
library(RColorBrewer)  
```

Extract and format data for nodulating 
```{r}
Venn <- read.xlsx("Venn.xlsx", sheet = 1)#Load your datafile
Nod.Positive <- Venn$Positive  # Extract "Positive" group 
Nod.Positive <- as.character(Nod.Positive) # Convert to character format (ensures compatibility for Venn diagram)
```

Extract and format data for non-nodulating 
```{r}
Nod.Negative <- Venn$Negative  # Extract "Negative" group 
Nod.Negative <- as.character(Venn$Negative)
```

Create a list containing both groups for Venn diagram plotting
```{r}
x <- list(Nod.Positive=Nod.Positive , Nod.Negative=Nod.Negative)

# Generate a Venn diagram showing only percentages
venn_plot <- ggvenn(x)

#Save the diagram in a PNG file
ggsave("venn_diagram.png", plot = venn_plot, width = 6, height = 6, dpi = 300)
```



